---
title: "Lab 4"
author: "Michael Velez"
date: "11:59PM March 10, 2021"
output:
  html_document:
    df_print: paged
---

Load up the famous iris dataset. We are going to do a different prediction problem. Imagine the only input x is Species and you are trying to predict y which is Petal.Length. A reasonable prediction is the average petal length within each Species. Prove that this is the OLS model by fitting an appropriate `lm` and then using the predict function to verify.

```{r}
data(iris)
mod = lm(Petal.Length ~ Species, iris)
mean(iris$Petal.Length[iris$Species == "setosa"])
mean(iris$Petal.Length[iris$Species == "versicolor"])
mean(iris$Petal.Length[iris$Species == "virginica"])
predict(mod, data.frame(Species = c("setosa")))
predict(mod, data.frame(Species = c("versicolor")))
predict(mod, data.frame(Species = c("virginica")))
```

Construct the design matrix with an intercept, $X$, without using `model.matrix`.

```{r}
X <- cbind(1, iris$Species == "versicolor", iris$Species == "virginica" )

head(X)

```

Find the hat matrix $H$ for this regression.

```{r}
H = X %*% solve(t(X) %*% X) %*% t(X)
Matrix::rankMatrix(H)

#head(H)
```

Verify this hat matrix is symmetric using the `expect_equal` function in the package `testthat`.

```{r}
pacman::p_load(testthat)
expect_equal(H, t(H))
```

Verify this hat matrix is idempotent using the `expect_equal` function in the package `testthat`.

```{r}
expect_equal(H, H%*%H)
```

Using the `diag` function, find the trace of the hat matrix.

```{r}
sum(diag(H))
#trace same as the rank
```

It turns out the trace of a hat matrix is the same as its rank! But we don't have time to prove these interesting and useful facts..

For masters students: create a matrix $X_\perp$.

```{r}
#TO-DO
```

Using the hat matrix, compute the $\hat{y}$ vector and using the projection onto the residual space, compute the $e$ vector and verify they are orthogonal to each other.

```{r}
y = iris$Petal.Length
y_hat = H %*% y
I = diag(nrow(iris))
e = (I-H) %*% y
e
```

Compute SST, SSR and SSE and $R^2$ and then show that SST = SSR + SSE.

```{r}
SSE = t(e) %*% e
SSE
y_bar = mean(y)
SST = t(y - y_bar) %*% (y - y_bar)
SST

Rsq = 1 - SSE/SST
Rsq

SSR = t(y_hat - y_bar) %*% (y_hat - y_bar)
SSR

expect_equal(SSR + SSE, SST)
```

Find the angle $\theta$ between $y$ - $\bar{y}1$ and $\hat{y} - \bar{y}1$ and then verify that its cosine squared is the same as the $R^2$ from the previous problem.

```{r}
theta = acos(t(y - y_bar) %*% (y_hat - y_bar) / sqrt(SST * SSR))
theta * (180 / pi)
```

Project the $y$ vector onto each column of the $X$ matrix and test if the sum of these projections is the same as yhat.

```{r}
proj1 = (X[,1] %*% t(X[,1]) / as.numeric(t(X[,1]) %*% X[,1])) %*% y 
proj2 = (X[,2] %*% t(X[,2]) / as.numeric(t(X[,2]) %*% X[,2])) %*% y 
proj3 = (X[,3] %*% t(X[,3]) / as.numeric(t(X[,3]) %*% X[,3])) %*% y 
```

Construct the design matrix without an intercept, $X$, without using `model.matrix`.

```{r}
X_2 = cbind(as.integer(iris$Species == "setosa"), as.integer(iris$Species == "versicolor"), as.integer(iris$Species == "virginica"))
head(X_2)
```

Find the OLS estimates using this design matrix. It should be the sample averages of the petal lengths within species.

```{r}
y = iris$Petal.Length
H = X_2 %*% solve(t(X_2) %*% X) %*% t(X_2)
y_hat = H %*% y
unique(y_hat)
```

Verify the hat matrix constructed from this design matrix is the same as the hat matrix constructed from the design matrix with the intercept. (Fact: orthogonal projection matrices are unique).

```{r}
X = cbind(as.integer(iris$Species == "setosa"), as.integer(iris$Species == "versicolor"), as.integer(iris$Species == "virginica"))
H = X_2 %*% solve(t(X_2) %*% X) %*% t(X_2)
H_new = X %*% solve(t(X) %*% X) %*% t(X)
expect_equal(H_new, H)
```

Project the $y$ vector onto each column of the $X$ matrix and test if the sum of these projections is the same as yhat.

```{r}
proj1 = ((X[,1] %*% t(X[,1])) / as.numeric(t(X[,1]) %*% X[,1])) %*% y
proj2 = ((X[,2] %*% t(X[,2])) / as.numeric(t(X[,2]) %*% X[,2])) %*% y
proj3 = ((X[,3] %*% t(X[,3])) / as.numeric(t(X[,3]) %*% X[,3])) %*% y
y_hat = H %*% y

expect_equal(proj1 + proj2 + proj3, y_hat)
```

Convert this design matrix into $Q$, an orthonormal matrix.

```{r}
Q = qr.Q(qr(X))
```

Project the $y$ vector onto each column of the $Q$ matrix and test if the sum of these projections is the same as yhat.

```{r}
proj1 = ((Q[,1] %*% t(Q[,1])) / as.numeric(t(Q[,1]) %*% Q[,1])) %*% y
proj2 = ((Q[,2] %*% t(Q[,2])) / as.numeric(t(Q[,2]) %*% Q[,2])) %*% y
proj3 = ((Q[,3] %*% t(Q[,3])) / as.numeric(t(Q[,3]) %*% Q[,3])) %*% y

expect_equal(proj1 + proj2 + proj3, y_hat)
```

Find the $p=3$ linear OLS estimates if $Q$ is used as the design matrix using the `lm` method. Is the OLS solution the same as the OLS solution for $X$?

```{r}
Q_mod = lm(Petal.Length ~ 0 + Q, iris)
Q_mod
```

Use the predict function and ensure that the predicted values are the same for both linear models: the one created with $X$ as its design matrix and the one created with $Q$ as its design matrix.

```{r}
colnames(X) = c("setosa", "versicolor", "virginica")
first_mod = lm(y ~ 0 + X)
unique(predict(first_mod, data.frame(X)))

second_mod = lm(y ~ 0 + Q)
unique(predict(second_mod, data.frame(Q)))
```


Clear the workspace and load the boston housing data and extract $X$ and $y$. The dimensions are $n=506$ and $p=13$. Create a matrix that is $(p + 1) \times (p + 1)$ full of NA's. Label the columns the same columns as X. Do not label the rows. For the first row, find the OLS estimate of the $y$ regressed on the first column only and put that in the first entry. For the second row, find the OLS estimates of the $y$ regressed on the first and second columns of $X$ only and put them in the first and second entries. For the third row, find the OLS estimates of the $y$ regressed on the first, second and third columns of $X$ only and put them in the first, second and third entries, etc. For the last row, fill it with the full OLS estimates.

```{r}
rm(list = ls())
boston = MASS::Boston
X = cbind(1, as.matrix(boston[,1:13]))
y = boston[,14]
p_add_one = ncol(X)

matrix_p_add_one = matrix(NA, nrow = p_add_one, ncol = p_add_one)
colnames(matrix_p_add_one) = c(colnames(boston[1:13]), "full OLS")
for (i in 1:ncol(X)) {
  X_i = X[,1:i]
  matrix_p_add_one[i,1:i] = solve(t(X_i) %*% X_i) %*% t(X_i) %*% y
}
matrix_p_add_one

```

Why are the estimates changing from row to row as you add in more predictors?

This is because each row is adding another feature which changes the estimates' value.

Create a vector of length $p+1$ and compute the R^2 values for each of the above models. 

```{r}
Rsq_vector = c(1:14)
for (i in 1:ncol(X)) {
  mod = lm(y ~ X[, 1:i])
  Rsq_vector[i] = summary(mod)$r.squared  
}
Rsq_vector
```

Is R^2 monotonically increasing? Why?

This is because as the amount of featurs goes up, the value of R^2 will increase (be more accurate).